Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-07-09 analysis/paper/paper.Rmd
fixed references, added figure captions, checked every chunks

Prior to European colonization, the indigenous people in northern Taiwan had already established their own exchange network that defined as "inter-insular trade" by @Chen2005 [p. 12] to refer to small-scale regional trade between China and Taiwan on irregular basis. This local exchange network of northeastern Taiwan had increasingly influenced by long distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since 15th century due to its location on the shipping routes, and was intensified in the 17th century when the arrival of the Europeans and their trade goods [@Chen2005; @Liu2017; @Wang2007]. Despite being on the periphery of major trade ports, Keelung and Tanshui, in northern Taiwan, Yilan was involved in this network through the periodic exchange activities between different indigenous groups. Yilan, also called "Kavalan" by local indigenous people since the prehistoric time, is a alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous people and another native group called Basai in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006].

@@

The documents written by Chinese historians in Qing dynasty usually described the decorative purposes of the ornaments and some documents had similar texts that might be copied from previous ones. According to those records, [@Chen2006[1852]: 308], [@Ko1993[1837]: 11, 126] and [@Yao1996[1829]: 77], it is mentioned that indigenous people in Yilan used ornaments in ceremonial contexts to display their status. Amoung those ornaments, golden fish-shaped necklaces made of copper threads had higher value due to its delicacy and more materials invested in production that usually possessed by rich people. Other people wore agate beads or glass beads on their head or neck to participating various ceremonies. In 1895, at the beginning of Japanese colonization, Ino Kanori [@Kanori1996, pp. 227-232] conducted a field survey for several aboriginal groups and found that golden necklaces had not been used in Yilan that time, but elder people still used beads.

Recent ethnographic research of the Kavalan tribe describes that agate beads were used in divination practice, called "subli", by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate bead is valuable object that are usually passed down from mother to daughter as a heirloom in shaman families without knowing the exact origin. Were not been using, beads would be often hided at invisible place such as on the roof beams to make sure no one sees that. During the Japanese colonization period, the number of agate beads declines due to the trade with Japanese who live in Taipei for collecting purpose [@Hsu1992, pp. 22] and golden fish-shaped necklaces had completely disappeared. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination that is consistent with the observation in European colonial period. It was usually the female shamans conducting divination ritual that passed down from their ancestors. Despite most aborigines today don’t know the origin of agate beads, the value of beads is still be highlighted through oral history and their scarcity. Those historical documents from the early 17th century to the modern ethnography described how local indigenous people use ornaments in local cultural contexts that represents some social roles or status. However, compared to European contact period, there is less mention of beads in Chinese contact period and the description about ornaments is only limited to their dressing culture.

@@

Kiwulan \@ref(fig:KWL-map) is located at northern Yilan and was a rescue archaeology projects that carried out from 2001 to 2004 in advance of construction. The total area includes eight open area sections that created 262 squares (4m by 4m) reaching 3,814 meter squares in total [@Chen2007]. Most ornaments were found in situ in the context while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The archaeological evidence includes a rich amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures that indicates it was a settlement. The chronology of Kiwulan can be divided into Lower Layer Culture (700 - 1200 AD) and Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between based on a series of 32 radiocarbon dates [@Chen2007 ]. This paper focuses on the Upper Layer Culture spanning from the late Iron Age that also called proto-historical period [@Liu2017], and historical period starting from the 17th century with European contact in Taiwan.

The earliest record of direct European contact with indigenous people in Yilan can be traced back to 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the Yilan Plain, with a population of 840 adults [@Nakamura1938, pp. 12]. The diagnostic artifacts of European contact found at Kiwulan including An-ping jars, stonewares that were largely introduced to Taiwan during the early 17th century. Those artifacts coupled with the original radiocarbon dates were used to identify the pre-European and post-European periods. The European colonization ended in 1662 when they were defeated by the kingdom of Tungning found by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The evidence of Chinese migration can be identified from Chinese official records and large amount of Chinese daily life porcelains found at Kiwulan.

@@

sample_size <- nrow(ornaments_period)
total_AD <- nrow(kwl_upper)
sample_pro <- sample_size/total_AD
```

@@

```{r, KWL-sam-area, fig.cap = "Sampling area with 40 squares", out.width= "400px", echo=FALSE}
# mapping sampling area
ggplot(sample_units) +
  geom_sf(data = AD_data, fill = NA) +
  geom_sf(color = 'red') +
  geom_sf_text(aes(label = `坑號`), alpha = 0.7) +
  theme_minimal() +
  theme_void() 
```

@@

```{r plot-all-ornaments-count, fig.cap= "Frequency of ornaments"}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories

@@

@@

Ornaments from the post-hole area accounts for the majority of the amount in total. Post-hole area provides a clear contexts and basis for comparison to see if there is any spatial pattern of distribution. For the post-hole area, there were `r sample_size` ornaments recovered from the 40 sampling squares, which accounts for `r sample_pro`% of the total number of ornaments found in the same open area section. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells. The variety of subtype and their individual frequency are summarized in \@ref(tab:tidy-data-for-table) in terms of three time episodes. [insert image here]

```{r plot-counts-by-period}
# categories of ornaments by period
ornaments_count <-
  ornaments_period %>% 
  filter(!is.na(`6-layers`),
         Categories != "Unknown Metal") %>%
  ungroup() %>% 
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  mutate(Categories = fct_infreq(Categories)) %>% 
  mutate(Categories = fct_rev(Categories)) %>% 
  group_by(period, Categories) %>%
  count()

ggplot(ornaments_count) +
  geom_col(aes(Categories, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(y = "Frequency", x = "Categories") +
  theme_minimal(base_size = 14)

# stack histogram
ggplot(ornaments_count) +
  geom_col(aes(Categories, n, fill = period)) +
  labs(y = "Frequency", x = "Categories") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  scale_fill_viridis_d()
```

@@

```{r plot-altogether, fig.cap= "Frequency of "}
library(cowplot)
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))
```
# Results

## Frequency between time periods
The results of the comparison between different time periods show a general pattern that most ornaments appeared in the pre-European contact period and the frequencies reached a peak during the European contact and then dropped during the Chinese contact period, especially for golden beads. This trend can be also seen on other ornaments including agate bead, metal ring, and bell. However, glass bead shows a different pattern that indicates a higher frequency in the pre-European contact, and then the frequency decreased in the European contact period and even less in the Chinese contact period.

The distribution of frequency for subtypes in each major category are presented in \@ref(fig:plot-all-ornaments-count). The distribution shows that agate beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European contact period. Agate beads can be divided into eight subtypes, in which hexagonal shape is the most common subtype that appeared before European contact and increased during the European contact and then declined in the Chinese contact. Small oval type shows similar frequency before and during European contact but not found in the Chinese contact period, while waxy oval bead only found during European contact period. There are more subtypes of agate beads during European contact compared to other time periods. Similarly metal rings can be classified into seven subtypes, in which wide small ring is the most common metal ornaments, followed by wide large and thin large shape. The greater varieties for those two categories might indicate the multiple sources due to global trade network brought by the Europeans. In contrast, copper bell and glass bead have smaller variety, but glass bead has larger number due to they were usually found in a cluster that reflect the original use as strings. Most common shape for copper bell is the large shape with human face as motif, while most common shape for glass bead is small that were identified as Indo-Pacific beads, the most common type that are widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

```{r plot-all-spatial-pattern, fig.cap= "Spatial pattern of ornaments"}
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join, 
          aes(geometry = geometry, 
              fill =n), 
          alpha = 0.9) +
  facet_wrap(~ period) +
  labs(fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))
```

```{r plot-spatial-pattern-each}
# filter golden bead
ornaments_period_join_gold <-
  ornaments_period_join %>% 
  filter(Categories %in% 'Golden bead')

# plot by period
ggplot(ornaments_period_join_gold) +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_gold, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_wrap(~ period) +
  labs(title = "Spatial Distribution of Golden Beads",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

@@

# plot by period
ggplot(ornaments_period_join_gold) +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_gold, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_wrap(~ period) +
  labs(title = "Spatial Distribution of Glass Beads",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

@@

# plot by period
ggplot(ornaments_period_join_gold) +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_gold, 
          aes(geometry = geometry, 
              fill = n), alpha = 0.9) +
  facet_wrap(~ period) +
  labs(title = "Spatial Distribution of Agate Beads",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

@@

# plot by period
ggplot(ornaments_period_join_metal) +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_metal, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_wrap(~ period) +
  labs(title = "Spatial Distribution of Metal Ring",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

@@

# plot by period
ggplot(ornaments_period_join_bell) +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_bell, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_wrap(~ period) +
  labs(title = "Spatial Distribution of Copper Bell",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))
```

## Frequency for spatial pattern
Figure \@ref(fig:plot-all-spatial-pattern) presents the spatial distribution of all classes of ornaments for each time period. It shows that one square unearthed more ornaments than others in the pre-European contact. During the European contact period, the presence of ornaments is more ubiquitous with some clusters on northern part, and more randomly distributed during the Chinese contact period. When looking at the distribution for each major class individually, it shows that some clusters of different class of ornaments across the area during the European contact period, including golden bead, agate bead, metal ring, and bell. However, it seems no clear consistent pattern of concentration for those ornaments. Each class shows its own pattern that squares with higher number of ornaments distributed separately and independently. For example, the cluster of golden bead were found at the northern part, while the cluster of agate bead were found in the middle part. In contrast, there are more than one cluster of metal rings that distribute separately across the research area. Copper bells were usually found solely and seem randomly across the area. Before European contact, the map shows greater amount of glass beads were found at the northern and middle part of the research area, and more widespread during the European contact period. In the Chinese Contact period, both the amount and density of different classes of ornaments decreases that they were only found in a few squares, for example, golden bead and glass bead were only found in a couple squares at the middle part.

@@

```{r plot-all-densities, fig.cap= "Kernal density map"}
par(mfrow = c(1, 3))
# apply function to each data frame in the list
titles <- names(ornaments_period_join_cen)

@@

## Spatial distribution
Point pattern analysis is a way to assess whether the distribution of ornaments presents some spatial pattern that can indicate the changes during culture contact period [@Bevan2016, @Ducke2015]. The location where ornaments from were square-based recorded, thus we took the intensity approach that focuses on the average density of points across space to explore their distribution. We randomly assigned points, here the location of ornaments, for each square according to the data, subset three groups for three time periods, and compute the kernal density for each time period for comparison. Kernal density estimations (KDE) can estimate the probability of the density of events across space by creating a continuous, smooth density surface across space. Here we use KDE to visualize the event of the presence of ornaments that were represents by maps on which we are able to see core areas and surrounding neighborhoods [@Bonnier2019, @Cortegoso2016]. Density value is assigned to each cell, of which the unit is meter. The results \@ref(fig:plot-all-densities) show that there is one major core area during the pre-European contact period, multiple core areas during European contact period, and a single core during the Chinese contact period. Those maps present three consistent sub-region with shifting core area over time. The distribution might indicate different social groups who own more ornaments, and the multiple groups during European contact period might reflect a more corporate mode of social organization. However, the generation of core areas might be biased due to small sample size, for example, a few ornaments found at one single square during the Chinese period could create obvious core area. Next step is to test whether the distribution of ornaments is randomly distributed to understand the underlying social indication related to ornaments.

@@

```{r hypo-test-distrubution, fig.cap= "Histogram of simulated ANN values"}

@@

Here we use Monte Carlo method to test for the complete spatial randomness of spatial events based on the average nearest-neighbor distance method. We hypothesize that the distribution of ornaments is consistent with a completely random process, and then simulate the presence of ornaments across the space for one thousand times for each time periods. The data was compared to the random pattern generated by simulated processes to see if the our data is randomly distributed or not. The results \@ref(fig:hypo-test-distrubution) are represented by histogram showing one thousand simulation with the blue line indicating our observed value. The results show that 100% of the simulated values are much more greater than our observed ANN value during the European contact period, which means the ornaments are clustered distributed. Similar result are also observed during the pre-European period but less extreme. The distribution of ornaments was random during the Chinese contact period and about one third of the simulated values are greater than our observed ANN value. This testing explains that the distribution of ornaments during the pre-European and European contact period might reflect the presence of different social groups. Moreover, the cluster of ornaments way more significant during the European contact period that might indicate some extent of control of ornaments.

@@

Q: Is the difference in the frequency of ornaments related to the intensity of contact, does it show any spatial pattern at Kiwulan that indicate the appearance of social inequality.

Q: Is the difference in the frequency of ornaments related to the scale of exchange network, does it show any spatial pattern at Kiwulan that indicate the appearance of social inequality.

>>>>

Prior to European colonization, the indigenous people in northern Taiwan had already established their own exchange network that defined as "inter-insular trade" by @Chen2005 [pp. 12] to refer to small-scale regional trade between China and Taiwan on irregular basis. This local exchange network of northeastern Taiwan had increasingly influenced by long distance trade between Fuzhou or Quanzhou in China and Ryukyu in Japan since 15th century due to its location on the shipping routes, and was intensified in the 17th century when the arrival of the Europeans and their trade goods [@Chen2005; @Liu2017; @Wang2007]. Despite being on the periphery of major trade ports, Keelung and Tanshui, in northern Taiwan, Yilan was involved in this network through the periodic exchange activities between different indigenous groups. Yilan, also called "Kavalan" by local indigenous people since the prehistoric time, is a alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous people and another native group called Basai in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006].

@@

The documents written by Chinese historians in Qing dynasty usually described the decorative purposes of the ornaments and some documents had similar texts that might be copied from previous ones without actual investigation. According to those records, indigenous people in Yilan had a custom to wear ornaments in ceremonial contexts to display their wealth and status [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. Among those ornaments, golden fish-shaped necklaces made of copper threads had higher value due to its delicacy and more materials invested in production that usually possessed by rich people. Other people wore agate beads or glass beads on their head or neck to participating various ceremonies. In 1895, at the beginning of Japanese colonization, @Ino2002 conducted a field survey for several aboriginal groups and found that golden necklaces had not been used in Yilan that time, but elder people still used beads.

Recent ethnographic research of the Kavalan tribe describes that agate beads were used in divination practice, called "subli", by female shamans [@Liu2008, pp. 133-134]. Interview records mention agate bead is valuable object that are usually passed down from mother to daughter as a heirloom in shaman families without knowing the exact origin. Were not been using, beads would be often hided at invisible place such as on the roof beams to make sure no one sees that. Modern ethnographic research shows that agate beads were not only used as decoration, but also played an important role in divination that is consistent with the observation in European colonial period. It was usually the female shamans conducting divination ritual that passed down from their ancestors. Despite most aborigines today don’t know the origin of agate beads, the value of beads is still be highlighted through oral history and their scarcity. Those historical documents from the early 17th century to the modern ethnography described how local indigenous people use ornaments in local cultural contexts that represents some social roles or status. However, compared to European contact period, there is less mention of beads in Chinese contact period and the description about ornaments is only limited to their dressing culture.

@@

Kiwulan \@ref(fig:KWL-map) is located at northern Yilan and was a rescue archaeology projects that carried out from 2001 to 2004 in advance of construction. The total area includes eight open area sections that created 262 squares (4m by 4m) reaching 3,814 meter squares in total [@Chen2007]. Most ornaments were found in situ in the context while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The archaeological evidence includes a rich amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures that indicates it was a settlement. The chronology of Kiwulan can be divided into Lower Layer Culture (700 - 1200 AD) and Upper Layer Culture (1400 - 1900 AD) with a sterile layer in between based on a series of 32 radiocarbon dates [@Chen2007]. This paper focuses on the Upper Layer Culture spanning from the late Iron Age that also called proto-historical period [@Liu2017], and historical period starting from the 17th century with European contact in Taiwan.

The earliest record of direct European contact with indigenous people in Yilan can be traced back to 1632, when the local villages were attacked by the Spanish who took revenge on an incident happened earlier that year [@Borao2001, pp. 163]. Later in 1647, the Dutch attacked the indigenous villages and forced them to accept colonial rules and economic demands by paying annual tribute [@Andrade2007]. According to Dutch census reports in 1650, Kiwulan was the largest indigenous settlement in the Yilan Plain, with a population of 840 adults [@Nakamura1938, pp. 12]. The diagnostic artifacts of European contact found at Kiwulan including An-ping jars, stonewares that were largely introduced to Taiwan during the early 17th century. Those artifacts coupled with the original radiocarbon dates were used to identify the pre-European and post-European periods. The European colonization ended in 1662 when they were defeated by the kingdom of Tungning found by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The evidence of Chinese migration can be identified both from Chinese official records and large amount of Chinese daily life porcelains found at Kiwulan [@Chen1963].

@@

sample_size <- nrow(ornaments_period)
total_AD <- nrow(kwl_upper)
sample_pro <- round(sample_size/total_AD, 1)
```

@@

```{r, KWL-sam-area, fig.cap = "Forty squares were sampled from the largest open area (in red with square ID number)", out.width= "70%", echo=FALSE}
# mapping sampling area
ggplot(sample_units) +
  geom_sf(data = AD_data, fill = NA) +
  geom_sf(color = 'red') +
  geom_sf_text(aes(label = `坑號`), size = 2.5, alpha = 0.9) +
  theme_minimal() +
  theme_void() 
```

@@

```{r plot-all-ornaments-count, fig.cap= "Frequency of ornaments", out.width='60%'}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories

@@

```{r, ornaments-image, fig.cap = "Subtypes of ornament in each major category. A: agate beads, B: bells, C: glass beads and golden beads, D: metal rings. Photographs are presented in the same order as those subtypes in the table but from left to right instead. The photographs of B, C, D categories were from original excavation report.", out.width= "400px", echo=FALSE}
knitr::include_graphics(here("analysis", "figures", "ornament.png"))
```

@@

Ornaments from the post-hole area accounts for the majority of the amount in total. Post-hole area provides a clear contexts and basis for comparison to see if there is any spatial pattern of distribution. For the post-hole area, there were `r sample_size` ornaments recovered from the 40 sampling squares, which accounts for `r sample_pro`% of the total number of ornaments found in the same open area section. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells. The variety of subtype and their individual frequency are summarized in \@ref(tab:tidy-data-for-table) in terms of three time episodes. The image for each subtype in major category shows in figure \@ref(fig:ornaments-image).

```{r, plot-counts-by-period, fig.cap = "Frequency of each category of ornament with a separate plot showing different time periods individually"}
# categories of ornaments by period
ornaments_count <-
  ornaments_period %>% 
  filter(!is.na(`6-layers`),
         Categories != "Unknown Metal") %>%
  ungroup() %>% 
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  mutate(Categories = fct_infreq(Categories)) %>% 
  mutate(Categories = fct_rev(Categories)) %>% 
  group_by(period, Categories) %>%
  count()

# histogram facetting for three periods
orna_periods_facet <-
ggplot(ornaments_count) +
  geom_col(aes(Categories, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(y = "Frequency") +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank())

# stack histogram
orna_all_viridis <-
ggplot(ornaments_count) +
  geom_col(aes(Categories, n, fill = period)) +
  labs(y = "Frequency", x = "Categories") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  scale_fill_viridis_d()

library(cowplot)
library(viridis)
# add two plots together
ggdraw() +
  draw_plot(orna_all_viridis + 
            theme(legend.justification = "top"), 0, 0, 1, 1) +
  draw_plot(orna_periods_facet + scale_color_viridis(discrete = TRUE) + 
              theme(legend.justification = "bottom"), 0.5, 0.15, 0.5, 0.5) +
  draw_plot_label(c("A", "B"), c(0, 0.5), c(1, 0.7), size = 15)
```

# Results

## Frequency between time periods
Figure \@ref(fig:plot-counts-by-period) shows the comparison between different time periods that indicates a general pattern that most ornaments existed already before European contact and the frequencies reached a peak during the European contact and then dropped during the Chinese contact period, especially for golden beads. This trend can be also seen on other ornaments including agate bead, metal ring, and bell. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then the frequency decreased in the European contact period and even much less in the Chinese contact period.

@@

```{r plot-altogether, fig.cap= "Frequency of "}
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))
```

The distribution of frequency for subtypes in each major category are presented in \@ref(fig:plot-all-ornaments-count). The distribution shows that agate beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European contact period. Agate beads can be divided into seven subtypes, in which hexagonal shape is the most common subtype that appeared before European contact and increased significantly during the European contact and then declined in the Chinese contact. Small oval type shows similar frequency before and during European contact but not found in the Chinese contact period, while waxy oval bead only found during European contact period. Higher variety of agate beads were found in European contact compared to other time periods. Similarly, metal rings can be classified into eight subtypes, in which wide small ring is the most common metal ornaments, followed by wide large and thin large shape. The greater varieties for those two categories might indicate the multiple origins due to global trade network stimulated by the Europeans. In contrast, copper bell and glass bead have less variety, but glass bead has larger number due to they were usually found in a cluster that indicates the original use as strings or necklaces. The common shape for copper bell is the large shape with a wide variety of human face as a motif, while the common shape for glass bead is bead with length less than 1cm that are identified as Indo-Pacific beads, the most common type that had been widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

```{r plot-all-spatial-pattern, fig.cap= "Spatial pattern of all class of ornament by time periods"}
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join, 
          aes(geometry = geometry, 
              fill =n), 
          alpha = 0.9) +
  facet_wrap(~ period) +
  labs(fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))
```

```{r, plot-spatial-pattern-each, fig.cap= "Spatial pattern for each class of ornament by time periods"}
# filter golden bead
ornaments_period_join_gold <-
  ornaments_period_join %>% 
  filter(Categories %in% 'Golden bead')

# plot by period
dis_golden <-
ggplot(ornaments_period_join_gold) +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_gold, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_wrap(~ period) +
  labs(title = "Golden Beads",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

@@

# plot by period
dis_glass <-
  ggplot(ornaments_period_join_gold) +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_gold, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_wrap(~ period) +
  labs(title = "Glass Beads",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

@@

# plot by period
dis_agate <-
ggplot(ornaments_period_join_gold) +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_gold, 
          aes(geometry = geometry, 
              fill = n), alpha = 0.9) +
  facet_wrap(~ period) +
  labs(title = "Agate Beads",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

@@

# plot by period
dis_metal <-
ggplot(ornaments_period_join_metal) +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_metal, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_wrap(~ period) +
  labs(title = "Metal Ring",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

@@

# plot by period
dis_bell <-
ggplot(ornaments_period_join_bell) +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_bell, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_wrap(~ period) +
  labs(title = "Copper Bell",
       fill = "frequency") +
  scale_fill_gradient(high = "black", low = "grey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

plot_grid(dis_golden,
          dis_glass,
          dis_agate, 
          dis_metal, 
          dis_bell,
          align = "v",
          nrow = 5)
```

## Frequency for spatial pattern
Figure \@ref(fig:plot-spatial-pattern-each) presents the spatial distribution of all categories for each time period. It shows that one square unearthed more ornaments than others in the pre-European contact. During the European contact period, ornaments was more ubiquitous with some clusters on northern part, and less presence during the Chinese contact period. When looking at the distribution for each major class individually, some clusters across the area were observed during the European contact including golden bead, agate bead, metal ring, and bell. However, it seems no clear consistent clustered pattern across those different ornaments. Each class shows its own pattern where the squares with higher number of ornaments distributed separately and independently. For example, the cluster of golden bead were found at the northern part, while the cluster of agate bead were found in the middle part. In contrast, there are more than one cluster of metal rings that distributed separately across the research area. Copper bells were usually found solely and seem randomly distributed across the area. Before European contact, the map shows greater amount of glass beads were found at the northern and middle part of the research area, and more widespread during the European contact period. In the Chinese Contact period, both the amount and density of different classes of ornaments decreased that they were only found in a few squares, for example, golden bead and glass bead were only found in a couple squares at the middle part.

@@

```{r plot-all-densities, fig.cap= "Kernal density map by periods"}
par(mfrow = c(1, 3))
# apply function to each data frame in the list
titles <- names(ornaments_period_join_cen)

@@

## Spatial distribution
To examine whether the distribution of ornament is uneven that could reflect the presence of social inequality, point pattern analysis is used to examine the distribution. Point pattern analysis is a way to assess whether the distribution of sites or artifacts represent some "hotspots" that imply social processes [@Bevan2016; @Ducke2015], and in this case the pattern of ornaments could be able to indicate the changes due to culture contact. The location where ornaments from were square-based recorded, thus the intensity approach was used that focuses on the average density of points across space to explore their distribution. Points were be randomly assigned to each square according to the frequency data. Next is to subset three groups for three time periods, and then compute the kernal density for each time period for comparison. Kernal density estimations (KDE) can estimate the probability of the density of events across space by creating a continuous, smooth density surface across space. Here we use KDE to visualize the event of the presence of ornaments that were represented by maps on which we are able to see core areas and surrounding neighborhoods [@Bonnier2019; @Cortegoso2016]. Density value is assigned to each cell, of which the unit is meter. The results \@ref(fig:plot-all-densities) show that there is one major core area during the pre-European contact period, multiple core areas during European contact period, and a single core during the Chinese contact period. Those maps present three consistent sub-region with shifting core area over time. The distribution might indicate different social groups who possessed more ornaments, and the multiple groups during European contact period might reflect some presence of unequal consumption. However, the generation of core areas might be biased due to small sample size, for example, a few ornaments found at one single square during the Chinese period could create obvious hot spot. To further evaluate the pattern, we test the hypothesis that whether the distribution of ornaments is randomly distributed to understand the underlying social indication related to ornaments.

@@

```{r hypo-test-distrubution, fig.cap= "Histogram of simulated ANN values with the blue line indicating the observed value"}

@@

Here we use Monte Carlo method to test the complete spatial randomness of spatial events based on the average nearest-neighbor distance method (ANN). We hypothesize that the distribution of ornaments is consistent with a completely random process, and then simulate the presence of ornaments across the space for one thousand times for each time periods. The data was compared to the random pattern generated by simulated processes to see if our data is randomly distributed or not. The results \@ref(fig:hypo-test-distrubution) are represented by histogram showing one thousand simulation. The results show that 100% of the simulated values are much more greater than our observed ANN value during the European contact period, which means the ornaments are clustered distributed. Similar result is also observed during the pre-European period but less extreme. The distribution of ornaments is more random during the Chinese contact period and about one third of the simulated values are greater than our observed ANN value. This testing explains that the distribution of ornaments during the pre-European and European contact period might reflect the presence of different social groups. Moreover, the cluster of ornaments way more significant during the European contact period that might indicate some extent of control of ornaments.

@@
